A scalar Poincaré map for anti-phase bursting in coupled inhibitory neurons with synaptic depression.

### 2022-03-05 tex/sections/introduction.tex
Edit introdu nullcline description

\documentclass[../manuscript.tex]{subfiles}
\graphicspath{{../figures/}}

\begin{document}
% DEPRESSION
Short-term synaptic plasticity may have a role in burst activity in central pattern
generators (CPGs).  Short-term synaptic depression is commonly found in neuronal networks
involved in the generation of rhythmic movements, such as in the pyloric CPG of the spiny
lobster~\citep{manor1997, rabbah2007}, or in the lumbosacral cord of the chick
embryo~\citep{donovan1998}.  Synaptic depression modulates the strength of synapses in
response to changes to the presynaptic firing frequency.  At a high neuronal firing
frequency, depression weakens the strength of synapses and therefore reduces the magnitude
of the postsynaptic response.  At low firing frequency, it allows sufficient time for the
synapse to recover from depression between spikes, leading to a stronger postsynaptic
response.  In reciprocal networks, synaptic depression has been shown to act as a
``switch'', giving rise to a wide range of network dynamics such as synchronous and
multi-stable rhythms, as well as fine tuning the frequency of network
oscillations~\citep{nadim2000, nadim1999, bose2011}.

@@

~\citet{bose2011} investigated burst generation in a generic half-centre CPG that consists
of two identical, tonically active Morris-Lecar~\citep{morris1981} neurons coupled through
inhibitory depressing synapses.  Numerical simulations showed that when the reciprocal
synaptic conductance between the two neurons is varied, the network produces symmetric
$n-n$ anti-phase bursts, with stronger synaptic coupling leading to longer bursts.  They
used methods from geometric singular perturbation theory to separate the timescales of the
fast membrane, and the slow synaptic dynamics of the network to derive one-dimensional
conditions necessary for the existence of stable $n-n$ solutions (for $n\leq 2$).
According to these conditions the type of firing pattern largely depends on the slow
depression dynamics of the synapses between the two neurons, and can therefore be
predicted by knowing the strengths of the synaptic conductances of the two synapses.
Thus, the scalar conditions derived in~\citet{bose2011} provide a method to numerically
identify the type of stable $n-n$ pattern for any given value of the coupling strength.
However, they do not predict the exact period of such solutions.  Furthermore, while they
provide good arguments for the validity of their reduction assumptions and the resulting
scalar conditions, they do not verify them numerically.

Here we extend the previous analysis by providing a Poincaré map of the slow depression
dynamics.  This allows us not only to predict the types of stable $n-n$ solutions the full
network can produce, (for any $n$), but also to study how varying the coupling strength
affects the period of such solutions.  To do this, we build on, and numerically test, the
assumptions on the fast-slow timescale disparity made in~\cite{bose2011}.  We reduce the
two-cell model to a scalar Poincaré map that tracks the evolution of the depression from
the beginning of one burst to the beginning of the next burst.  Stable fixed points of our
map are associated with stable $n-n$ burst solutions.  Our map construction is motivated
by the burst length map of a T-type calcium current, utilised by~\citet{matveev2007},
which approximates the anti-phase bursting dynamics of a network of two coupled
Morris-Lecar neurons.  In contrast to our model, the network described in
the~\cite{matveev2007} paper does not contain short-term synaptic depression, and burst
termination is instead accomplished through the dynamics of a slow T-type calcium current.

% MAIN RESULTS AND CONCLUSIONS
The Poincaré map derived here replicates the results from numerical simulations of the
full two-cell ODE system: Given the strength of maximum conductance between the two
neurons, fixed points of our map predict the type and period of $n-n$ patterns, the switch
between burst solutions of different periods, as well as the occurrence of co-existent
solutions.  In addition to proving the existence and stability of fixed points, our map
shows that fixed points are created via a fold bifurcation of maps.  We also derive
algebraic conditions that predict the period increment bifurcations that allow the switch
between $n-n$ and $(n+1)-(n+1)$ solutions.  Because our map is fully explicit, it lays the
framework for studying the effects of other model parameter on network dynamics without
the need to run expensive numerical integrations of the ODEs.

% PAPER STRUCTURE.
This paper is organised as follows.  First, we introduce the network of two neurons, and
describe the properties of single cell and synapse dynamics.  We use numerical simulations
of the network to provide an intuition for the range of possible burst dynamics the system
can produce.  Next, we state and justify the simplifying assumptions that are necessary
for the map construction.  Finally, we analytically derive the first return map of the
depression variable as well as the conditions that are required for stable solutions to
bifurcate between $n-n$ and $(n+1)-(n+1)$ solutions.  We end this work with a discussion.

\end{document}

>>>>

\section{Introduction}
% DEPRESSION
Short-term synaptic plasticity may have a role in burst activity in central pattern
generators (CPGs).  Short-term synaptic depression is commonly found in neuronal networks
involved in the generation of rhythmic movements, such as in the pyloric CPG of the spiny
lobster~\citep{manor1997, rabbah2007}, or in the lumbosacral cord of the chick
embryo~\citep{donovan1998}.  Synaptic depression modulates the strength of synapses in
response to changes to the presynaptic firing frequency.  At a high neuronal firing
frequency, depression weakens the strength of synapses and therefore reduces the magnitude
of the postsynaptic response.  At low firing frequency, it allows sufficient time for the
synapse to recover from depression between spikes, leading to a stronger postsynaptic
response.  In reciprocal networks, synaptic depression has been shown to act as a
``switch'', giving rise to a wide range of network dynamics such as synchronous and
multi-stable rhythms, as well as fine tuning the frequency of network
oscillations~\citep{nadim2000, nadim1999, bose2011}.

@@

~\citet{bose2011} investigated burst generation in a generic half-centre CPG that consists
of two identical, tonically active Morris-Lecar~\citep{morris1981} neurons coupled through
inhibitory depressing synapses.  Numerical simulations showed that when the reciprocal
synaptic conductance between the two neurons is varied, the network produces symmetric
\(n-n\) anti-phase bursts, with stronger synaptic coupling leading to longer bursts.  They
used methods from geometric singular perturbation theory to separate the timescales of the
fast membrane, and the slow synaptic dynamics of the network to derive one-dimensional
conditions necessary for the existence of stable \(n-n\) solutions (for \(n\leq 2\)).
According to these conditions the type of firing pattern largely depends on the slow
depression dynamics of the synapses between the two neurons, and can therefore be
predicted by knowing the strengths of the synaptic conductances of the two synapses.
Thus, the scalar conditions derived in~\citet{bose2011} provide a method to numerically
identify the type of stable \(n-n\) pattern for any given value of the coupling strength.
However, they do not predict the exact period of such solutions.  Furthermore, while they
provide good arguments for the validity of their reduction assumptions and the resulting
scalar conditions, they do not verify them numerically.

Here we extend the previous analysis by providing a Poincaré map of the slow depression
dynamics.  This allows us not only to predict the types of stable \(n-n\) solutions the full
network can produce, (for any \(n\)), but also to study how varying the coupling strength
affects the period of such solutions.  To do this, we build on, and numerically test, the
assumptions on the fast-slow timescale disparity made in~\cite{bose2011}.  We reduce the
two-cell model to a scalar Poincaré map that tracks the evolution of the depression from
the beginning of one burst to the beginning of the next burst.  Stable fixed points of our
map are associated with stable \(n-n\) burst solutions.  Our map construction is motivated
by the burst length map of a T-type calcium current, utilised by~\citet{matveev2007},
which approximates the anti-phase bursting dynamics of a network of two coupled
Morris-Lecar neurons.  In contrast to our model, the network described in
the~\cite{matveev2007} paper does not contain short-term synaptic depression, and burst
termination is instead accomplished through the dynamics of a slow T-type calcium current.

% MAIN RESULTS AND CONCLUSIONS
The Poincaré map derived here replicates the results from numerical simulations of the
full two-cell ODE system: Given the strength of maximum conductance between the two
neurons, fixed points of our map predict the type and period of \(n-n\) patterns, the switch
between burst solutions of different periods, as well as the occurrence of co-existent
solutions.  In addition to proving the existence and stability of fixed points, our map
shows that fixed points are created via a fold bifurcation of maps.  We also derive
algebraic conditions that predict the period increment bifurcations that allow the switch
between \(n-n\) and \((n+1)-(n+1)\) solutions.  Because our map is fully explicit, it lays the
framework for studying the effects of other model parameter on network dynamics without
the need to run expensive numerical integrations of the ODEs.

% PAPER STRUCTURE.
This paper is organised as follows.  First, we introduce the network of two neurons, and
describe the properties of single cell and synapse dynamics.  We use numerical simulations
of the network to provide an intuition for the range of possible burst dynamics the system
can produce.  Next, we state and justify the simplifying assumptions that are necessary
for the map construction.  Finally, we analytically derive the first return map of the
depression variable as well as the conditions that are required for stable solutions to
bifurcate between \(n-n\) and \((n+1)-(n+1)\) solutions.  We end this work with a discussion.

### 2022-03-05 tex/sections/methods.tex
Edit introdu nullcline description

\documentclass[../manuscript.tex]{subfiles}
\graphicspath{{../figures/}}

\begin{document}
We consider a pair of identical Morris-Lecar neurons~\citep{morris1981}, with parameters
adapted from~\cite{bose2011}.  The Morris-Lecar model is a set of two first-order
differential equations that describe the membrane dynamics of a spiking neuron.  The
depolarisation is modelled by an instantaneous calcium current, and the hyperpolarisation
by a slow potassium current and a leak current.  The membrane potential $v_{i}$ and
potassium activation $w_{i}$ of neuron $i$ ($i, j=1,2$) is described by:
\begin{align}
  ~\label{eq:cell-modelA}
  \dot v_{i} & = f(v_{i}, w_{i}) -\gbar s_j(v_i-v_{s}), \\
  ~\label{eq:cell-modelB}
  \dot w_{i} & =h(v_i,w_i).
\end{align}
Here $v_{s}$ is the inhibitory reversal potential, and $\gbar$ and $s_{j}$ are the maximal
synaptic conductance and the synaptic gating, respectively, constituting the total
inhibitory conductance $\gbar s_{j}$ from neuron $j$ to neuron $i$.  Function $f(v_{i},
w_{i})$ describes the membrane currents of a single cell:
\begin{equation}
  ~\label{eq:memcur}
  f(v_{i}, w_{i}) = -g_{\ca}m_{\infty}(v_{i})(v_{i}-v_{\ca}) - g_{\k}w_{i}(v_{i}-v_{\k})-g_{\leak}(v_{i}-v_{\leak}) + I.
\end{equation}
The currents include a constant current $I$, and three ionic currents: an instantaneous
calcium current, a potassium current, and a leak current, with respective reversal
potentials $v_{\ca}$, $v_{\k}$, and $v_{\leak}$, as well as maximum conductances
$g_{\ca}$, $g_{\k}$, and $g_{\leak}$.  The function $h(v_{i}, w_{i})$ models the kinetics
of the potassium gating variable $w_{i}$, and is given by
\begin{equation}
  h(v_{i}, w_{i})=\frac{w_{\infty}(v_{i})-w_{i}}{\tau_{w}}~\label{eq:h}.
\end{equation}
The steady-state activation functions $m_{\infty}$ and $w_{\infty}$ as well as the default
model parameters are described in the \textbf{Supplementary Material S1}.

The dynamics of the synaptic interactions between the neurons are governed by a synaptic
gating variable $s_{i}$ and a depression variable $d_{i}$:
\begin{linenomath}
  \begin{align}
    \dot d_{i} & =
    \begin{cases}
      (1-d_{i})/\tau_a\text{ if }v_i<v_\theta, \\
      -d_{i}/\tau_b\text{ if }v_i>v_\theta,
    \end{cases}\label{eq:dot-d} \\
    \dot s_{i} & =
    \begin{cases}
      -s_i/\tau_\kappa\text{ if }v_i<v_\theta, \\
      (d_i-s_i)/\tau_\gamma\text{ if }v_i>v_\theta.
    \end{cases}\label{eq:dot-s}
  \end{align}
\end{linenomath}
Variable $d_{i}$ describes a firing rate dependent depletion mechanism that governs the amount of depression acting on the synapse.
The model is agnostic with respect to the exact mechanism of this depletion, be it pre- or post-synaptic.
When the voltage is below firing threshold, depression variable $d_i$ recovers with time constant $\tau_a$, while synaptic variable $s_i$ decays with time constant $\tau_\kappa$.
% TODO Maybe mention that this topic will be discussed in the appendix or later.
Because synaptic depression occurs on a much slower timescale than synaptic inhibition, we assume $\tau_d\gg\tau_\kappa$.
When the voltage is above firing threshold, variable $d_i$ decays to zero with $\tau_b$, while $s_i$ quickly approaches the respective value of $d_i$.
Since $\tau_\gamma$ is the shortest time constant of the system with $\tau_\gamma \ll 1$, we can assume that whenever $v>v_\theta$ we can we have $s_i=d_i$.
The equations for the depression model were adapted from the~\citet{bose2001} model.
These equations are a mathematically tractable simplification of the established phenomenological depression model previously described by~\citet{tsodyks1997}.

When the cells are uncoupled ($\gbar=0$), the membrane dynamics are determined by the cubic $v$-nullcline $v_{\infty}(v_i)$ and the sigmoid $w$-nullcline $w_{\infty}(v_{i})$, satisfying $\dot v_{i}=0$ and $\dot w_{i}=0$, respectively.
The two curves intersect along the middle branch of $v_{\infty}$, creating an unstable fixed point $p_{f}=(v_{f},w_{f})$ with a surrounding stable limit cycle of period $T=T_{act}+T_{inact}$ (\cref{fig:nullclines}A).
Here $T_{act}$ is the amount of time the cell spends in the active state when $v>v_\theta$, while $T_{inact}$ is the time it spends in the silent state when $v<v_\theta$.
Trajectories along that limit cycle have the familiar shape of the action potential (\cref{fig:nullclines}B).
The trajectory of an action potential can be dissected into four phases: (1) a silent phase, (2) a jump up, (3) an active phase, and (4) a jump down~\citep[see e.g.][]{ermentrout2010}.
During the silent phase the trajectory evolves along the left branch ($v_{i}<v_{\theta}$) of the cubic $v$-nullcline.
Once the trajectory reaches the local minimum of $v_{\infty}$, it ``jumps up'' to the right branch ($v_{i}>v_{\theta}$), crossing the firing threshold $v_{\theta}$.
During the active phase the trajectory then evolves along the right branch of the cubic until it arrives at the local maximum, where it ``jumps down'' to the left branch commencing a new cycle.

The two-cell network model is numerically integrated using an adaptive step-size integrator for stiff differential equations implemented with XPPAUT~\citep{ermentrout2002} and controlled through the Python packages SciPy~\citep{scipy2020} and PyXPP~\citep{pyxpp}.
The following mathematical analysis is performed on the equations of a single cell.
Unless required for clarity, we will therefore omit the subscripts $i,j$ from here on.

\end{document}

>>>>

\section{Materials and Methods}
We consider a pair of identical Morris-Lecar neurons \citep{morris1981}, with parameters adapted from \cite{bose2011}.  The Morris-Lecar model is a set of two first-order differential equations that describe the membrane dynamics of a spiking neuron.  The depolarisation is modelled by an instantaneous calcium current, and the hyperpolarisation by a slow potassium current and a leak current.  The membrane potential \(v_{i}\) and potassium activation \(w_{i}\) of neuron \(i\) (\(i, j=1,2\)) is described by: \begin{align}
	\label{eq:cell-modelA}
	\dot v_{i} & = f(v_{i}, w_{i}) -\gbar s_j(v_i-v_{s}), \\
	\label{eq:cell-modelB}
	\dot w_{i} & =h(v_i,w_i).
\end{align}
Here \(v_{s}\) is the inhibitory reversal potential, and \(\gbar\) and \(s_{j}\) are the
maximal synaptic conductance and the synaptic gating, respectively, constituting the total
inhibitory conductance \(\gbar s_{j}\) from neuron \(j\) to neuron \(i\).  Function
\(f(v_{i}, w_{i})\) describes the membrane currents of a single cell:
\begin{equation}
	\label{eq:memcur}
	f(v_{i}, w_{i}) = -g_{\ca}m_{\infty}(v_{i})(v_{i}-v_{\ca}) - g_{\k}w_{i}(v_{i}-v_{\k})
	-g_{\leak}(v_{i}-v_{\leak}) + I.
\end{equation}
The currents include a constant current \(I\), and three ionic currents: an instantaneous
calcium current, a potassium current, and a leak current, with respective reversal
potentials \(v_{\ca}\), \(v_{\k}\), and \(v_{\leak}\), as well as maximum conductances
\(g_{\ca}\), \(g_{\k}\), and \(g_{\leak}\).  The function \(h(v_{i}, w_{i})\) models the
kinetics of the potassium gating variable \(w_{i}\), and is given by
\begin{equation}
	h(v_{i}, w_{i})=\frac{w_{\infty}(v_{i})-w_{i}}{\tau_{w}}~\label{eq:h}.
\end{equation}
The steady-state activation functions \(m_{\infty}\) and \(w_{\infty}\) as well as the default
model parameters are described in the \textbf{Supplementary Material S1}.

The dynamics of the synaptic interactions between the neurons are governed by a synaptic
gating variable \(s_{i}\) and a depression variable \(d_{i}\):
\begin{numcases}{\dot d_{i} = }
	(1-d_{i})/\tau_{a} & \text{ if }v_{i}<v_{\theta},\label{eq:dot-d-up}
	\\
	-d_{i}/\tau_{b}    & \text{ if }v_{i}>v_{\theta},\label{eq:dot-d-down}
\end{numcases}

\begin{numcases}{\dot s_{i} = }
	-s_{i} / \tau_\kappa & \text{ if }v_{i}<v_{\theta},\label{eq:dot-s-down}
	\\
	0                & \text{ if }v_{i}>v_{\theta.}\label{eq:dot-s-up}
\end{numcases}

% TODO: Needs EDITING
Variable \(d_{i}\) describes a firing rate dependent depletion mechanism that governs the amount of depression acting on the synapse. The model is agnostic with respect to the exact mechanism of this depletion, be it pre- or post-synaptic. When the voltage of cell \(i\) is above firing threshold (\(v_i>v_\theta\)), variable \(d_i\) decays with time constant \(\tau_b\), and recovers with time constant \(\tau_a\) when voltage is below firing threshold (\(v_i < v_\theta\)). Since the synaptic inhibition occurs on a much faster timescale than synaptic depression, we assume that \(s_i\) is instantaneously reset to \(d_i\) whenever \(v_i\) increases above \(v_\theta\), where it remains throughout \(v_i > v_\theta\). Whenever \(v_i < v_\theta\), the synaptic variable decays exponentially with time constant \(\tau_\kappa\). The equations for the depression model were adapted from the \citet{bose2001} model. These equations are a mathematically tractable simplification of the established phenomenological depression model previously described by \citet{tsodyks1997}.

When the cells are uncoupled (\(\gbar=0\)), the membrane dynamics are determined by the cubic \(v\)-nullcline \(v_{\infty}(v_i)\) and the sigmoid \(w\)-nullcline \(w_{\infty}(v_{i})\), satisfying \(\dot v_{i}=0\) and \(\dot w_{i}=0\), respectively.
The two curves intersect along the middle branch of \(v_{\infty}\), creating an unstable fixed point with a surrounding stable limit cycle of period \(T=T_{act}+T_{inact}\) (\cref{fig:nullclines}A).
Here \(T_{act}\) is the amount of time the cell spends in the active state when \(v>v_\theta\), while \(T_{inact}\) is the time it spends in the silent state when \(v<v_\theta\).
Trajectories along that limit cycle have the familiar shape of the action potential (\cref{fig:nullclines}B).
The trajectory of an action potential can be dissected into four phases: (1) a silent phase, (2) a jump up, (3) an active phase, and (4) a jump down~\citep[see e.g.][]{ermentrout2010}.
During the silent phase the trajectory evolves along the left branch (\(v_{i}<v_{\theta}\)) of the cubic \(v\)-nullcline.
Once the trajectory reaches the local minimum of \(v_{\infty}\), it ``jumps up'' to the right branch (\(v_{i}>v_{\theta}\)), crossing the firing threshold \(v_{\theta}\).
During the active phase the trajectory then evolves along the right branch of the cubic until it arrives at the local maximum, where it ``jumps down'' to the left branch commencing a new cycle.

The two-cell network model is numerically integrated using an adaptive step-size integrator for stiff differential equations implemented with XPPAUT~\citep{ermentrout2002} and controlled through the Python packages SciPy~\citep{scipy2020} and PyXPP~\citep{pyxpp}. The following mathematical analysis is performed on the equations of a single cell. Unless required for clarity, we will therefore omit the subscripts \(i,j\) from here on.
